### 2025-04-18 multiple-regression.qmd
20250417 - nonlinearity and heteroscedasticity

To evaluate the shape of the association between the [predictor variables](#sec-correlationalStudy) and the [outcome variable](#sec-correlationalStudy), we can examine [scatterplots](#sec-scatterplot), residual plots (@fig-residualPlots), marginal model plots (@fig-marginalModelPlots), added-variable plots (@fig-addedVariablePlots), and component-plus-residual plots (@fig-componentPlusResidualPlots).
Residual plots depict the residuals (errors) on the y-axis as a function of the fitted values or a specific [predictor](#sec-correlationalStudy) on the x-axis.
Marginal model plots are basically glorified [scatterplots](#sec-scatterplot) that depict the [outcome variable](#sec-correlationalStudy) (both in terms of observed values and model-fitted values) on the y-axis and the [predictor variables](#sec-correlationalStudy) on the x-axis.
Added-variable plots depict the *unique* association of each [predictor variables](#sec-correlationalStudy) with the [outcome variable](#sec-correlationalStudy) when controlling for all the other [predictor variables](#sec-correlationalStudy) in the model.
Component-plus-residual plots depict partial residuals on the y-axis as a function of each [predictor variable](#sec-correlationalStudy) on the x-axis, where a partial residual for a given [predictor](#sec-correlationalStudy) is the effect of a given predictor (thus controlling for all the other [predictor variables](#sec-correlationalStudy) in the model) plus the residual from the full model.

@@

To evaluate homoscedasticity, we can evaluate a residual plot (@fig-residualPlots) and spread-level plot (@fig-spreadLevelPlot).
A residual plot depicts the residuals on the y-axis as a function of the model's fitted values on the x-axis.
Homoscedasticity in a residual plot is identified as a constant spread of residuals versus fitted values—the residuals do not show a fan or cone shape; a fan or cone shape indicates heteroscedasticity.
A spread-level plot depicts the log of the absolute value of studentized residuals on the y-axis as a function of the log of the model's fitted values on the x-axis.
Homoscedasticity in a spread-level plot is identified as a flat slope; a slope that differs from zero indicates heteroscedasticity.

>>>>

To evaluate the shape of the association between the [predictor variables](#sec-correlationalStudy) and the [outcome variable](#sec-correlationalStudy), we can examine [scatterplots](#sec-scatterplot) (@fig-linearNonlinearAssociations), residual plots (@fig-residualPlots), marginal model plots (@fig-marginalModelPlots), added-variable plots (@fig-addedVariablePlots), and component-plus-residual plots (@fig-componentPlusResidualPlots).
Residual plots depict the residuals (errors) on the y-axis as a function of the fitted values or a specific [predictor](#sec-correlationalStudy) on the x-axis.
Marginal model plots are basically glorified [scatterplots](#sec-scatterplot) that depict the [outcome variable](#sec-correlationalStudy) (both in terms of observed values and model-fitted values) on the y-axis and the [predictor variables](#sec-correlationalStudy) on the x-axis.
Added-variable plots depict the *unique* association of each [predictor variables](#sec-correlationalStudy) with the [outcome variable](#sec-correlationalStudy) when controlling for all the other [predictor variables](#sec-correlationalStudy) in the model.
Component-plus-residual plots depict partial residuals on the y-axis as a function of each [predictor variable](#sec-correlationalStudy) on the x-axis, where a partial residual for a given [predictor](#sec-correlationalStudy) is the effect of a given predictor (thus controlling for all the other [predictor variables](#sec-correlationalStudy) in the model) plus the residual from the full model.

Examples of linear and nonlinear associations are depicted with [scatterplots](#sec-scatterplot) in @fig-linearNonlinearAssociations.

```{r}
#| layout-ncol: 2
#| label: fig-linearNonlinearAssociations
#| fig-cap: "Example Associations Depicted With Scatterplots."
#| fig-alt: "Example Associations Depicted With Scatterplots."
#| fig-subcap:
#|   - "Example of a Linear Association"
#|   - "Example of a Nonlinear Association"
#| code-fold: true

sampleSize <- 1000
quadraticX <- runif(
  sampleSize,
  min = -4,
  max = 4)
linearY <- quadraticX + rnorm(sampleSize, mean = 0, sd = 0.5)
quadraticY <- quadraticX ^ 2 + rnorm(sampleSize)
quadraticData <- cbind(quadraticX, quadraticY) %>%
  data.frame %>%
  arrange(quadraticX)

plot(
  x = quadraticData$quadraticX,
  y = quadraticData$quadraticY,
  xlab = "",
  ylab = "",
  main = "Nonlinear Association")

plot(
  x = quadraticX,
  y = linearY,
  xlab = "",
  ylab = "",
  main = "Linear Association")

abline(
  lm(linearY ~ quadraticX),
  lwd = 2,
  col = "blue")
```

@@

To evaluate homoscedasticity, we can evaluate a residual plot (@fig-residualPlots) and spread-level plot (@fig-spreadLevelPlot).
A residual plot depicts the residuals on the y-axis as a function of the model's fitted values on the x-axis.
Homoscedasticity in a residual plot is identified as a constant spread of residuals versus fitted values—the residuals do not show a fan, cone, or bow-tie shape; a fan, cone, or bow-tie shape indicates heteroscedasticity.
In a residual plot, a fan or cone shape indicates increasing or decreasing variance in the residuals as a function of the fitted values; a bow-tie shape indicates that the residuals are smallest in the middle of the fitted values and greatest on the extremes of the fitted values.
A spread-level plot depicts the log of the absolute value of studentized residuals on the y-axis as a function of the log of the model's fitted values on the x-axis.
Homoscedasticity in a spread-level plot is identified as a flat slope; a slope that differs from zero indicates heteroscedasticity.

```{r}
#| layout: [[100], [50, 50]]
#| label: fig-homscedasticityHeteroscedasticity
#| fig-cap: "Example of Homoscedasticity and Heteroscedasticity in Residual Plots."
#| fig-alt: "Example of Homoscedasticity and Heteroscedasticity in Residual Plots."
#| fig-subcap:
#|   - "Homoscedasticity"
#|   - "Fan-Shaped Heteroscedasticity"
#|   - "Bow-Tie-Shaped Heteroscedasticity"
#| code-fold: true

# 1. Homoscedasticity
sampleSize <- 1000
x1 <- runif(sampleSize, 0, 10)
y1 <- 3 + 2 * x1 + rnorm(sampleSize, mean = 0, sd = 2)
fit1 <- lm(y1 ~ x1)
res1 <- resid(fit1)
fitted1 <- fitted(fit1)

# 2. Fan-shaped heteroscedasticity
x2 <- runif(sampleSize, 0, 10)
y2 <- 3 + 2 * x2 + rnorm(sampleSize, mean = 0, sd = 0.5 * x2) # increasing variance
fit2 <- lm(y2 ~ x2)
res2 <- resid(fit2)
fitted2 <- fitted(fit2)

# 3. Bow-tie-shaped heteroscedasticity
x3 <- runif(sampleSize, 0, 10)
sd3 <- abs(x3 - 5) + 0.5  # variance smallest in middle, higher on edges
y3 <- 3 + 2 * x3 + rnorm(sampleSize, mean = 0, sd = sd3)
fit3 <- lm(y3 ~ x3)
res3 <- resid(fit3)
fitted3 <- fitted(fit3)

plot(
  fitted1,
  res1,
  main = "Homoscedasticity",
  xlab = "Fitted value",
  ylab = "Residual")

abline(
  h = 0,
  lwd = 2,
  col = "blue")

plot(
  fitted2,
  res2,
  main = "Fan-Shaped Heteroscedasticity",
  xlab = "Fitted value",
  ylab = "Residual")

abline(
  h = 0,
  lwd = 2,
  col = "blue")

plot(
  fitted3,
  res3,
  main = "Bow-Tie-Shaped Heteroscedasticity",
  xlab = "Fitted value",
  ylab = "Residual")

abline(
  h = 0,
  lwd = 2,
  col = "blue")
```
